---
title: "Water Table EDA"
author: "Zach Turner"
date: "September 10, 2016"
output: html_document
---

This notebook will be used to explore the Water Table dataset.

### Training Data
The training data comes in two parts, one is the independent variables and the second is the dependent variable. I'll combine those together right away before doing anything else.

```{r,message=FALSE,warning=FALSE}
library(dplyr)
library(ggplot2)
library(cowplot)
setwd("~/Git/Contests/DrivenData/WaterTable")
training1 <- read.csv("training.csv",sep=",")
training2 <- read.csv("training_labels.csv",sep=",")
training <- inner_join(training1,training2,by="id")
rm(training1,training2)
```

### Basics

First I want to see a few things about the dataset as a whole, and then I'll break a bit deeper into looking at the numeric/character variables.

```{r}
dim(training)

# Return any variables with missing values
colnames(training)[lapply(training,function(x) mean(is.na(x)))>0]

# Target
ggplot(data=training) + geom_bar(aes(x=status_group,y=..count..),fill="dodgerblue") + ggtitle("Target Distribution")

```

So I don't have any missing data, which is great. I can also see that most of the pumps are either functional or not functional. Pumps that need repair are more rare and will probably be a bit tougher to identify. 

```{r}
table(sapply(training,class))
```

We have 31 factor variables (including the response), and then 10 numeric predictors. 

### Numeric Data  

Since I don't have too many numeric variables, it won't be too tough to explore them individually. I want to examine their histograms and then their correlations with the target.

```{r}
numvars <- colnames(training)[sapply(training,class)!="factor"]
numvars
```

#### Histograms

I want to examine some histograms broken apart by their target group to see how the distributions may differ.

```{r}
training %>% filter(amount_tsh < 10000) %>% ggplot(aes(x=amount_tsh)) + geom_histogram(aes(y=..count../sum(..count..)),bins=50) + facet_wrap(~status_group) + labs(ylab("Proportion"))
```

So across the board, amount_tsh is mostly 0. (Amount_tsh refers to the height of the water discharge location relative to the water in the tank)

```{r}
training %>% ggplot(aes(x=gps_height)) + geom_histogram(aes(y=..count../sum(..count..)),bins=50) + facet_wrap(~status_group) + labs(ylab("Proportion"))
```

We can sort of see the same thing with the gps_height, which is the altitude of the well (most are at sea level).

I also want to see if I can carve out any details from the latitude/longitude.

```{r}
training %>% filter(longitude>0) %>% ggplot(aes(x=longitude,fill=as.factor(status_group))) + geom_histogram(position="fill",binwidth=0.3) + scale_fill_brewer(palette="Blues") + labs(fill="Status Group")
```

```{r}
training %>% filter(latitude < -1) %>% ggplot(aes(x=latitude,fill=as.factor(status_group))) + geom_histogram(position="fill",binwidth=0.3) + scale_fill_brewer(palette="Blues") + labs(fill="Status Group")
```

There appear to be some regions where the pumps are more functional. Probably the most interesting find is that longitudes near 30 tend to have more pumps that need repair -- this may come in handy.

The vast majority of num_private is 0, so I want to split this out separately.
```{r}
training %>% filter(num_private==0) %>% ggplot() + geom_bar(aes(x=num_private,fill=status_group),position = "fill")

training %>% filter(num_private !=0,num_private < 1000) %>% ggplot() + geom_histogram(aes(x=num_private),bins=50) + facet_wrap(~status_group)
```

This doesn't really seem to tell us anything new.

```{r}
training %>% filter(population < 1000) %>% ggplot() + geom_histogram(aes(population),fill="dodgerblue")
```

I would expect to see that the older pumps are failing more often.

```{r}
training %>% filter(construction_year>0) %>% ggplot() + geom_histogram(aes(x=construction_year,fill=status_group),bins=54)
```

Ok, this is good. The older pumps tend to be less functional. However, we only have the ages for roughly half of the observations.